---
title: "Developer Guide"
author:
  - "Adriana Godamunne"
  - "Kavya R. Kathiravan"
  - "Braden Fuhriman"
  - "Connor J. Bell"
  - "Matthew J. Madison"
  - "Michael E. Cotterell"
output: rmarkdown::html_vignette
description: >
  A comprehensive resource for contributing to TDCM.
vignette: >
  %\VignetteIndexEntry{Developer Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
) # knitr::opts_chunk$set
```

```{r, include = FALSE}
# determine the current package maintainer
person_pattern <- "(?<name>.*?) <(?<email>.*?)>"
tdcm_pkg <- utils::packageDescription("TDCM")
tdcm_maintainer <- stringr::str_match(tdcm_pkg$Maintainer, person_pattern)[1, ]
tdcm_maintainer_name <- tdcm_maintainer["name"]
tdcm_maintainer_email <- tdcm_maintainer["email"]
```

::: {.pkgdown-release}
<div class="alert alert-warning small" role="alert">
A more recent version of this page may be available
[here](https://cotterell.github.io/tdcm/dev/articles/developer.html).
</div>
:::

This guide is a comprehensive resource for contributing to the _R_ package 
**TDCM** and is intended for both new and experienced contributors. 
Contributions to TDCM are always welcome; however, to keep things consistent and
easier to maintain, the package maintainer asks that all contributors follow the
policies and guidelines included in this document when preparing and submitting 
their contributions to the package.

## Types of Contributions

**Would you like to help contribute to TDCM?** We are more than happy to work 
with our community to improve our user experience! Some ways that you or your
team can contribute are listed below. You can click on each contribution type
to learn more. 

* [Issue Reporting](#issue-reporting): Report any bugs or issues you come across 
  so that we can try to fix them
  
* [Feature Requests](#feature-requests): Let us know about features you would 
  like to see added to the package.
  
* [Code Contributions](#code-contributions): Help us work on the package by 
  fixing bugs or adding features yourself!

### Issue Reporting

If you notice any errors, bugs, or unexpected behaviors in the TDCM package, 
please use the Issues feature on GitHub. This will allow us to respond promptly 
and keep all required fixes in one place. To do so, log in to your GitHub 
account (or create one if you do not already have one), then follow the 
instructions below:

1. Navigate to our [issues page](https://github.com/cotterell/tdcm/issues) on 
   GitHub.
   
2. Use the search area to see if the same issue you encountered or something 
   similar has already been reported.
   
   * If an existing issue report applies to your specific issue, then open that
     report and read the comment history. If you are unable to find a solution
     in the comments, then leave a new comment to let us know that you have
     encountered the same issue or something similar along with enough details
     for a developer to reproduce your version of the issue. One of the first 
     things the development team will do is attempt to reproduce the issue, so 
     these additional details are very important and including them often leads
     to quicker fixes.
     
   * If the issue has not been reported yet, submit a new issue report:
          
     1. Click the green "New Issue" button.
     2. Click "Get Started” next to "Bug Report"
     3. Click "Submit New Issue” once you have finished filling out the form.
     
3. Whether you commented on an existing issue report or submitted a new one, you 
   can track the progress of the issue report by visiting its page on GitHub.
   If the developers need reach out to you for more information, they will do so
   by mentioning your name in a new comment that they add to the report. When
   this happens, GitHub should notify you about it. Replying to the developer
   should help the developers resolve the issue more quickly!

### Feature Request

**Have any features you would like us to add?** Please let us know using the 
Issues feature on GitHub!

> *Issues* is a feature on GitHub that streamlines communication between users 
> and developers and consolidates all project tasks in one place. While this of 
> course includes bugs, it is often used for managing feature requests and 
> improvements, allowing it to serve as a "To-Do List" developers can use to 
> know what they should be focusing on.

To create an Issue report, log in to your GitHub account (or create one if you 
do not already have one) and follow the instructions below:

1. Navigate to our [issues page](https://github.com/cotterell/tdcm/issues) on
   GitHub.
   
2. Use the search box to see if the feature you’re thinking of has already been 
   requested.
	 
	 * If there is already a request, please interact with it!
	 
	 * You can use the emoji icon to add a reaction, such as a "thumbs up" to show
	   your support! This will let us know which features to prioritize by showing 
	   how many people want a particular feature. 
	 
	 * Additionally, feel free to voice your opinions in the replies of the 
	   request page! You can add additional details you’d like to see implemented,
	   or simply express your desire for the feature to be added!
	 
	 * If there is no existing request for the feature, you can make your own:
	 
	   1. Click the green "New Issue" button
	   2. Click "Get Started” next to "Feature Request"
	   3. Click "Submit New Issue” once you’ve finished filling out the form!
	   
3. Once you have posted your feature request, developers may reply to your 
   request asking for more details. Please provide as much information as 
   possible so the developers know what you are looking for.

### Code Contributions

For your contribution to be accepted, it needs to follow a consistent style, be 
well documented, and include unit tests. You can ensure your code meets our 
standards by following the steps outlined below. All package instructions are 
written for use in RStudio.

To install all necessary packages, run the following commands:

TODO finish this section.

## Git & Pull Requests

The below section describes what Git is and how to use it to help contribute to this repository.

### What is Git?

[Git](https://git-scm.com/book/en/v2/Getting-Started-What-is-Git%3F) is a 
version control system that allows developers to track changes in their code 
over time. It's like a time machine for your project files, enabling you to save
different versions (called "commits") of your code. Each commit records what 
changes were made, by whom, and when, making it easy to revert back to older 
versions if something goes wrong with the current code.

This is a breakdown of key features and concepts in Git:

* [Repository](https://git-scm.com/book/en/v2/Git-Basics-Getting-a-Git-Repository) (or *repo*): 
  This is where all your project’s files and their history are stored. It can be 
  located on your computer (local repository) or on a remote server (remote 
  repository).
  
* [Commit](https://git-scm.com/book/en/v2/Git-Basics-Recording-Changes-to-the-Repository): 
  A commit is like a snapshot of your project at a specific point in time. When 
  you make changes to your files, you 'commit' these changes, which saves a 
  record of what was altered, allowing you to track the evolution of your 
  project.
  
* [Branch](https://git-scm.com/book/en/v2/Git-Basics-Recording-Changes-to-the-Repository): 
  Branches allow you to create separate versions of your project within the same
  repository. You might use a branch to develop a new feature while keeping the 
  main project (usually called the "master" or "main" branch) stable.
  
* [Merge](https://git-scm.com/book/en/v2/Git-Basics-Recording-Changes-to-the-Repository): 
  When you've finished working in a branch, you can merge your changes back into 
  the main branch. This combines the histories of both branches, integrating the
  new features or changes into the main project.
  
* [Pull](https://git-scm.com/book/en/v2/GitHub-Maintaining-a-Project): Pulling 
  refers to downloading changes from a remote repository to your local one. This
  is useful when working in teams, as it lets you see and integrate the changes 
  made by others.
  
* [Push](https://git-scm.com/docs/git-push): Pushing refers to uploading your 
  local repository changes to a remote repository. This is how you share your 
  work with others or back up your current state to a remote location.

Using [Git](https://git-scm.com/video/what-is-git), 
[multiple](https://git-scm.com/video/what-is-version-control) developers can 
work on the same project simultaneously without interfering with each others' 
work, as they can work in different branches and merge changes only when ready. 
This collaborative feature, along with the ability to track every change, makes
Git an essential tool for modern software development.

### Git vs. GitHub

[Git](https://git-scm.com/) is a version control system (VCS) that helps 
you track changes in your code files. It allows multiple developers to work 
together on the same project effectively and efficiently by providing tools to
manage differing versions of code. Git is used to create branches, handle 
merges, revert to previous states, and more, directly from your local machine 
or a server without needing an internet connection. Git operates primarily on 
your local system, managing your local repository. However, it can synchronize 
with remote repositories.

[GitHub](https://github.com/) is a hosting service for Git 
repositories. It is a platform where you can upload your Git repositories and 
collaborate with others. Beyond the basic functionality of Git, GitHub 
provides additional features such as bug tracking, feature requests, task 
management, and wikis for project documentation. GitHub is designed to 
facilitate easier collaboration among team members on a project. It offers 
graphical tools for managing pull requests, code reviews, and even project 
management tools like issues and project boards.
  
In essence, Git is the tool that enables version control, while GitHub is a 
service that hosts your repositories and adds extra features to enhance 
collaboration among team members. Both work hand-in-hand to support the 
development process, but they serve different roles within that framework.

Further Reading:

* Chacon, S., & Straub, B. (November 2014).
  "[Getting Started: What is Git?](https://git-scm.com/book/en/v2/Getting-Started-What-is-Git%3F)"
  _Pro Git_. Second Edition. Apress. ISBN: 978-1484200773.

### What is a Pull Request?

A [pull request](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/reviewing-changes-in-pull-requests/about-pull-request-reviews) 
(PR), also known as a [merge request](https://git-scm.com/book/en/v2/Git-Branching-Basic-Branching-and-Merging) 
in some systems like GitLab, is a feature used in Git repositories that lets 
developers notify team members about changes they've pushed to a branch in a 
repository on platforms like GitHub, Bitbucket, or GitLab. Essentially, it's a 
request to review and then merge this branch into another branch, typically the 
main project branch.

Here's how it works in simple terms:

* [Branch Creation](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-a-pull-request): 
  First, a developer creates a branch from the main code base to work on a new 
  feature, a bug fix, or any updates. This separation ensures that the main 
  code base remains unaffected until the new changes are reviewed and approved.
  
* [Making Changes and Committing](https://git-scm.com/book/en/v2/Git-Basics-Recording-Changes-to-the-Repository): 
  The developer makes necessary changes in this new branch, commits these 
  changes, and pushes the branch to a remote repository.
  
* Opening a Pull Request: Once the branch is updated on the remote repository, 
  the developer opens a pull request. This PR is essentially a request to the 
  rest of the team to review the changes made in the branch before it gets 
  merged into the main branch.

* Review and Discussion: Team members review the changes, discuss potential 
  improvements, and may request further 
  [changes](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/about-comparing-branches-in-pull-requests) 
  if needed. This process is crucial for maintaining code quality and ensuring 
  that the integration aligns with the project's overall direction.

* Approval and [Merge](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/resolving-a-merge-conflict-on-github): 
  After the changes are reviewed and approved by the necessary team members or 
  according to the project's guidelines, the pull request can be merged. This 
  means that the changes in the branch will be integrated into the target 
  branch, usually the main or master branch.
  
* Closure and Cleanup: Once merged, the pull request is closed, and it's common 
  to delete the branch to keep the repository clean and manageable.
  
Pull Requests are a vital part of the collaborative development process, 
enabling transparency, improving code quality, and helping manage contributions 
to a project efficiently. They facilitate a peer review system which is 
beneficial for both team cohesion and project integrity.

### Using RStudio to Prepare a Pull Request

RStudio does not have a built-in method of creating a pull request. However, it 
has a built-in terminal that can be used to push changes to a forked repository 
with a GitHub remote origin. GitHub must then be used to create a pull request.

1. On GitHub: Click the “Fork” button on the project to make changes to.

2. On RStudio: Create a new project and use Version Control with Git. Copy and 
   paste the repository URL of your fork of the project.
   
3. Add the original repository as an upstream version. This should be done in 
   RStudio’s built-in terminal. Type in:
   
   ```
   git remote add upstream https://github.com/cotterell/tdcm.git
   ```
   
4. Create and checkout a branch in which you will make changes.

5. Commit your changes using the “Git” tab on RStudio. Ensure that you are on 
   the branch with your changes.
   
6. Push to your forked repository in the terminal: 

   ```
   git push origin <branch name>
   ```
   
7. On GitHub: Click the green icon in the top left of your repository page and 
   create a pull request with a message describing your changes.

### Using Visual Studio Code to Prepare a Pull Request

TODO write.

### Using a Terminal to Prepare a Pull Request

You can use [GitHub’s CLI (downloadable)](https://cli.github.com) to fork a 
repository and create a pull request within it. Fork the original project using
[`gh repo fork <repository>`](https://cli.github.com/manual/gh_repo_fork), then
commit your changes on a separate branch. Once all of your changes are
committed, create a pull request using 
[`gh pr create`](https://cli.github.com/manual/gh_pr_create) 

#### Continuous Integration and Pull Requests

This package utilizes continuous integration via GitHub Actions. This means that
a script within the project will automatically run various **checks** on a pull 
request and notify you of any problems on your **GitHub profile**. Once these 
are resolved, provided that they do not require further inspection, your pull 
request will be reviewed by the development team for potential inclusion into
the package.

## Policies

TODO write. For example, do not modify the `DESCRIPTION` file unless you are the
package maintainer.

## Guidelines

TODO write.

## Maintainer Notes

This section includes notes that are intended for the package maintainer only. 
At the time of this writing, the package maintainer is `r tdcm_maintainer_name` 
based on the package's `DESCRIPTION` file.

### GitHub Actions

#### R-CMD-check

[![R-CMD-check](https://github.com/cotterell/tdcm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/cotterell/tdcm/actions/workflows/R-CMD-check.yaml)

This workflow runs `R-CMD-check` using the three major operating systems (linux, 
macOS and Windows) with the current, development, and previous versions of R.

#### pkgdown

[![pkgdown](https://github.com/cotterell/tdcm/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/cotterell/tdcm/actions/workflows/pkgdown.yaml)

This workflow builds the package website using 
`pkgdown::build_site_github_pages()`, then pushes the built package website to 
GitHub Pages. The inclusion of
[`workflow_dispatch`](https://docs.github.com/en/actions/learn-github-actions/events-that-trigger-workflows#workflow_dispatch)
means that the workflow can 
[be run manually](https://docs.github.com/en/actions/managing-workflow-runs/manually-running-a-workflow)
or [triggered via the GitHub REST API](https://docs.github.com/en/rest/reference/actions/#create-a-workflow-dispatch-event).

- Release website: [https://cotterell.github.io/tdcm/](https://cotterell.github.io/tdcm/)
- Development website: [https://cotterell.github.io/tdcm/dev/](https://cotterell.github.io/tdcm/dev/)
  
#### pr-commands

[![pr-commands](https://github.com/cotterell/tdcm/actions/workflows/pr-commands.yaml/badge.svg)](https://github.com/cotterell/tdcm/actions/workflows/pr-commands.yaml) 

This workflow adds `/document` and `/style` commands for pull requests.

#### test-coverage

[![test-coverage](https://github.com/cotterell/tdcm/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/cotterell/tdcm/actions/workflows/test-coverage.yaml)

This workflow uses `covr::codecov()` to query the test coverage of the package and upload the result 
to [https://app.codecov.io/gh/cotterell/tdcm](https://app.codecov.io/gh/cotterell/tdcm).
 
### Prepare a Release

TODO write.

## References

Chacon, S., & Straub, B. (November 2014). 
_Pro Git_. 
Second Edition. Apress. ISBN: 978-1484200773.
[https://git-scm.com/book/en/v2](https://git-scm.com/book/en/v2)

Wickham, H., & Bryan, J. (July 2023). 
_R Packages: Organize, Test, Document, and Share Your Code._ 
Second Edition. O'Reilly Media. ISBN: 9781098134945.
[https://r-pkgs.org/](https://r-pkgs.org/)

Wickham, H., Hester, J., Chang. W., & Bryan, J. (2022). 
_devtools: Tools to Make Developing R Packages Easier_. 
[https://devtools.r-lib.org/](https://devtools.r-lib.org/), 
[https://github.com/r-lib/devtools](https://github.com/r-lib/devtools)
